\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-10-08 NAR_DRAFT.tex
hpc acknowledgement
new supplementary figure
Update title & abstract

\title{\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation}

@@

\begin{abstract}
%The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and identification of lower frequency variants while providing well calibrated false positive rate expectations.
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}

>>>>

\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}

@@

\begin{abstract}
Identifying somatic mutations withins tumors is key to understanding treatment resistance, patient prognosis, and tumor evolution.
Mutations at low allelic frequency, which are carried by only a portion of cells within a tumor, are particularly difficult to detect.
Many algorithms have been developed to identify such mutations, but none models a key aspect of tumor biology.
Namely, every tumor has its own profile of mutation types that it tends to generate.
We present \batcave (Bayesian Analysis Tools for Context-Aware Variant Evaluation), an algorithm that first learns the tumor mutational profile and then uses it as a prior for evaluating potential mutations.
We also present an R implementation of the algorithm that is built on the popular somatic variant caller MuTect.
Using simulations, we show that adding the \batcave algorithm to MuTect improves variant identification.
It also improves the calibration of posterior probabilities, enabling a more principled tradeoff between precision and recall.
Lastly, we show that \batcave also performs well on real data.
Our implementation is computationally inexpensive and straightforward to incorporate into existing MuTect pipelines.
More broadly, the algorithm can be added to other variant callers, and it can be extend to include additional biological features that affect mutation generation.
\end{abstract}
